---
title: "Case Study 2 EDA and Linear Regression Model"
author: "Jack Korbitz"
date: "2024-11-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load data and packages
```{r}
wine = read.csv("/Users/jackkorbitz/School/MSDS_6306/Unit 14 and 15/Wine Train Set.csv", header = TRUE)
wine = wine |>
  mutate(location = ifelse(location == "Califormia", "California", location))
winetest = read.csv("/Users/jackkorbitz/School/MSDS_6306/Unit 14 and 15/Wine Test Set.csv", header = TRUE)
winetest = winetest |>
  mutate(location = ifelse(location == "Califormia", "California", location))

library(tidyverse)
library(plotly)
library(GGally)
library(ggthemes)
library(corrplot)
library(knitr)
library(reshape2)
library(forecast)
library(stats)

```

# Initial Data Exploration

```{r}
## Preview the data
head(wine)
summary(wine)
selected_vars = wine[, c("fixed.acidity", "volatile.acidity", "residual.sugar", "pH", "alcohol", "quality")]

# Check if 'quality' exists and convert it to a factor for coloring
if ("quality" %in% colnames(wine)) {
  selected_vars$quality <- as.factor(selected_vars$quality)
}

# Visualize pairwise relationships with 'quality' as color
ggpairs(
  selected_vars,
  columns = 2:5, 
  aes(color = quality, alpha = .8),
  title = "Pairwise Relationships with Quality as Color",
  lower = list(continuous = "smooth"),
  upper = list(continuous = "cor"))
```

# Initial EDA of the data
## Just to better understand how the data is structured and to see if I can pull out any patterns
```{r}
# fixed.acidity
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = fixed.acidity, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Fixed Acidity by Quality",
    x = "Quality",
    y = "Fixed Acidity",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# volatile.acidity
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = volatile.acidity, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Volatile Acidity by Quality",
    x = "Quality",
    y = "Volatile Acidity",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# citric.acid
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = citric.acid, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Citric Acid by Quality",
    x = "Quality",
    y = "Citric Acid",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# residual.sugar
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = residual.sugar, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Residual Sugar by Quality",
    x = "Quality",
    y = "Residual Sugar",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
 # ggplotly(p)
  
# chlorides
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = chlorides, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Chlorides by Quality",
    x = "Quality",
    y = "Chlorides",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# free.sulfur.dioxide
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = free.sulfur.dioxide, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Free Sulfar Dioxide by Quality",
    x = "Quality",
    y = "Free Sulfar Dioxide",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# total.sulfur.dioxide
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = total.sulfur.dioxide, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Total Sulfar Dioxide by Quality",
    x = "Quality",
    y = "Total Sulfar Dioxide",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# density
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = density, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Density by Quality",
    x = "Quality",
    y = "Density",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  ggplotly(p)
  
# pH
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = pH, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of pH by Quality",
    x = "Quality",
    y = "pH",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# sulphates
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = sulphates, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Sulphates by Quality",
    x = "Quality",
    y = "Sulphates",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
  
# alcohol
#p = 
  ggplot(wine, aes(x = as.factor(quality), y = alcohol, fill = as.factor(quality))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Alcohol by Quality",
    x = "Quality",
    y = "Alcohol",
    fill = "Quality"
  ) +
  facet_grid(type ~ location) +
  theme_wsj() 
  #ggplotly(p)
```

# EDA by Type
## Break out insight by Red vs White wine
```{r}
# Create scatterplot
ggplot(wine, aes(x = density, y = residual.sugar, color = type)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Density vs. Residual Sugar",
    x = "Density",
    y = "Residual Sugar",
    color = "Type"
  ) +
  theme_solarized()


ggplot(wine, aes(x = total.sulfur.dioxide, y = residual.sugar, color = type)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Total SO2 vs. Residual Sugar",
    x = "Total Sulfur Dioxide",
    y = "Residual Sugar",
    color = "Type"
  ) +
  theme_solarized()

ggplot(wine, aes(x = total.sulfur.dioxide, y = volatile.acidity, color = type)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Total SO2 vs. Volatile Acidity",
    x = "Total Sulfur Dioxide",
    y = "Volatile Acidity",
    color = "Type"
  ) +
  theme_solarized()

ggplot(wine, aes(x = density, y = alcohol, color = type)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Density vs. Alcohol",
    x = "Density",
    y = "Alcohol",
    color = "Type"
  ) +
  theme_solarized()

ggplot(wine, aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide, color = type)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Total SO2 vs. Free SO2",
    x = "Free SO2",
    y = "Total SO",
    color = "Type"
  ) +
  theme_solarized()


# Create the 3D scatter plot for Type
plot_ly(data = wine, 
        x = ~density, 
        y = ~total.sulfur.dioxide,
        z = ~volatile.acidity, 
        type = "scatter3d", 
        mode = "markers",
        color = ~as.factor(type), 
        colors = c("#C8102E","#0033A0"), 
        marker = list(size = 4, opacity = 0.7)) %>%
  layout(title = "3D Scatterplot: Type of Wine by Density, Total SO2, and Volatile Acidity",
         scene = list(xaxis = list(title = "Density"),
                      yaxis = list(title = "Total SO2"),
                      zaxis = list(title = "Volatile Acidity")),
         legend = list(title = list(text = "Legend")))
```

# EDA by Location
## Break out insight by California vs Texas
```{r}
ggpairs(
  wine,
  columns = c('fixed.acidity', 'volatile.acidity', 'citric.acid', 'residual.sugar', 'chlorides', 'free.sulfur.dioxide', 'total.sulfur.dioxide', 'density', 'pH', 'sulphates', 'alcohol','quality'),
  aes(color = location, alpha = .8),  
  title = "Pairwise Relationships with Quality as Color",
  lower = list(continuous = "smooth"),
  upper = list(continuous = "cor")
)


# Calculate mean quality and count by location
summary_table <- wine %>%
  group_by(location) %>%
  summarise(
    count = n(),
    mean_quality = mean(quality, na.rm = TRUE)
  )

# Display the table using kable
kable(summary_table, format = "pipe", col.names = c("Location", "Count", "Mean Quality"))


ggplot(wine, aes(x = as.factor(quality), fill = location)) +
  geom_bar(position = "fill", color = "black", width = 0.7) + 
  geom_text(stat = "count", aes(label = ..count..), 
            position = position_fill(vjust = 0.5), size = 3, color = "white", fontface = "bold") +  
  labs(
    title = "Quality by Location: California vs. Texas",
    x = "Wine Quality",
    y = "Proportion",
    fill = "Location"
      ) +
  scale_fill_manual(values = c("California" = "#0033A0", "Texas" = "#C8102E")) +  
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title = element_text(size = 12),  
    axis.text = element_text(size = 10),   
    legend.position = "top" 
       )


ggplot(wine, aes(x = alcohol, y = volatile.acidity, color = location)) +
  geom_point(alpha = .6) +
  labs(
    title = "Scatterplot of Factor 1 vs. Factor 2",
    x = "Factor 1",
    y = "Factor 2",
    color = "Location"
  ) +
  theme_solarized()


plot_ly(data = wine, 
        x = ~alcohol, 
        y = ~density,
        z = ~volatile.acidity, 
        type = "scatter3d", 
        mode = "markers",
        color = ~location, 
        colors = c("#0033A0", "#C8102E"),
        marker = list(size = 4, opacity = 0.5)) |>
  layout(title = "3D Scatterplot: Location of Wine by Alcohol, Density, and Volatile Acidity",
         scene = list(xaxis = list(title = "Alcohol"),
                      yaxis = list(title = "Density"),
                      zaxis = list(title = "Volatile Acidity")),
         legend = list(title = list(text = "Legend")))
```

# EDA by Quality
## 3D viz on quality picked from CM
```{r}
# create a 3D scatterplot for quality
plot_ly(data = wine, 
        x = ~density, 
        y = ~volatile.acidity,
        z = ~alcohol, 
        type = "scatter3d", 
        mode = "markers",
        color = ~as.factor(quality),
        colors = c("#FF0000",  "#FF5733", "#FFFF33","#33FF57","#33FFF6","#0099ff", "#9900ff"),
        marker = list(size = 4,  opacity = 0.8)) |>
  layout(title = "3D Scatterplot: Location of Wine by Volatile Acidity, Density, and Alcohol",
         scene = list(xaxis = list(title = "Volatile Acidity"),
                      yaxis = list(title = "Density"),
                      zaxis = list(title = "Alcohol")),
         legend = list(title = list(text = "Legend")))
```
# Create a correlation matrix to select variables for the Linear Regression Model

```{r}
# Encode 'location' and 'type' as a numeric values
wine$type <- as.numeric(as.factor(wine$type))
wine$location <- as.numeric(as.factor(wine$location))
numeric_columns <- wine[, sapply(wine, is.numeric)]
# Compute the correlation matrix
cor_matrix <- cor(numeric_columns, use = "complete.obs")
# Visualize the correlation matrix with numbers
library(corrplot)
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix Including Location and Type", mar = c(0, 0, 1, 0),
         addCoef.col = "black", 
         number.cex = 0.7,
         col = colorRampPalette(c("#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC"))(200))
```

# Linear Regression Model
## Create a linear regression model to predict wine quality
```{r}
wine$quality <- as.integer(wine$quality)

# Train/test split (80/20)
set.seed(123456)
split <- 0.80
train_indices <- sample(seq(1, nrow(wine)), round(split * nrow(wine)))
wine_train_model <- wine[train_indices, ]
wine_test_model  <- wine[-train_indices, ]

# Create the variable
predictors <- c(
                "alcohol"
               ,"chlorides"
               ,"citric.acid"
               ,"density"
               ,"fixed.acidity"
               ,"free.sulfur.dioxide"
               ,"pH"
               ,"residual.sugar"
               ,"sulphates"
               ,"total.sulfur.dioxide"
               ,"volatile.acidity"
                )

# Calculate mean and standard deviation for scaling
scaling_params <- data.frame(
  mean = colMeans(wine_train_model[predictors]),
  sd = apply(wine_train_model[predictors], 2, sd)
)

# Scale training data
wine_train_model[predictors] <- scale(wine_train_model[predictors])

# Scale test data using training set statistics
wine_test_model[predictors] <- sweep(wine_test_model[predictors], 2, scaling_params$mean, "-")
wine_test_model[predictors] <- sweep(wine_test_model[predictors], 2, scaling_params$sd, "/")

# Function to calculate MAE
calculate_mae <- function(model, model_name, wine_test_model) {
  # Make predictions on the test set
  pred <- predict(model, newdata = wine_test_model)
  
  # Calculate MAE (Mean Absolute Error)
  mae <- mean(abs(pred - wine_test_model$quality))
  
  # Return the model name and MAE
  return(data.frame(model = model_name, mae = mae))
}

# Initialize an empty data frame to store MAE results
mae_results <- data.frame(model = character(), mae = numeric())

# Linear Regression Model
# Dynamically create the formula for predictors
formula <- as.formula(paste("quality ~", paste(predictors, collapse = " + ")))
lm_model <- lm(formula, data = wine_train_model)

# Calculate MAE for Linear Regression
lm_mae <- calculate_mae(lm_model, "Linear Regression", wine_test_model)
mae_results <- rbind(mae_results, lm_mae)

# Print results
print(mae_results)

```

# Use the model to predict the quality of the test set
```{r}
# Ensure 'quality' is an integer (if it's not already)
wine$quality <- as.integer(wine$quality)

# Train/test split (80/20)
set.seed(123456)
split <- 0.80
train_indices <- sample(seq(1, nrow(wine)), round(split * nrow(wine)))
wine_train_model <- wine[train_indices, ]
wine_test_model  <- wine[-train_indices, ]

# Predictor variables
predictors <- c(
                ,"alcohol"
                ,"chlorides"
                ,"citric.acid"
                ,"density"
                ,"fixed.acidity"
                ,"free.sulfur.dioxide"
                ,"pH"
                ,"residual.sugar"
                ,"sulphates"
                ,"total.sulfur.dioxide"
                ,"volatile.acidity"
                )

# Calculate mean and standard deviation for scaling
scaling_params <- data.frame(
  mean = colMeans(wine_train_model[predictors]),
  sd = apply(wine_train_model[predictors], 2, sd)
)

# Scale training data
wine_train_model[predictors] <- scale(wine_train_model[predictors])

# Build the linear regression model
formula <- as.formula(paste("quality ~", paste(predictors, collapse = " + ")))
lm_model <- lm(formula, data = wine_train_model)

# Scale the winetest data using training set scaling parameters
winetest_scaled <- winetest
winetest_scaled[predictors] <- sweep(winetest_scaled[predictors], 2, scaling_params$mean, "-")
winetest_scaled[predictors] <- sweep(winetest_scaled[predictors], 2, scaling_params$sd, "/")

# Predict wine quality on the scaled winetest data
winetest_scaled$quality <- round(predict(lm_model, newdata = winetest_scaled))  # Round predictions to integers

# Create the output dataframe with only `id` and `quality`
output <- winetest_scaled %>% select(ID, quality)

# Save the output to a user-specified file path
save_predictions <- function(output, file_path) {
  write.csv(output, file = file_path, row.names = FALSE)
  cat("Predictions saved to:", file_path, "\n")
}

# Example usage: drop the desired file path as a string
file_path <- "/Users/jackkorbitz/School/MSDS_6306/Unit 14 and 15/Predict_Quality.csv"
save_predictions(output, file_path)

```